---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview: 

In this hands-on exercise, I learn how to import and wrangling geospatial data using appropriate R packages.

## Getting Started

The code chunk below install and load [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://tidyverse.tidyverse.org/) packages into R environment.

Firstly, ensure you have downloaded the pacman library, else run install.packages("pacman") on your console before proceeding with the below code.

```{r}
pacman::p_load(sf, tidyverse)

```

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data

```{r}
cyclingpath = st_read(dsn = "data/geospatial",
                      layer = "CyclingPath")
```

### Importing GIS data in kml format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

## Checking the Content of A Simple Feature Data Frame

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

### Working with glimpse()

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

```{r}
plot(mpsz)
```

Using max.plot = 15 to plot all attributes;

```{r}
# To visualise the geospatial features
plot(mpsz, max.plot = 15)
```

```{r}
# To plot only the geometry
plot(st_geometry(mpsz))
```

```{r}
# To plot the sf object by using specific attribute
plot(mpsz["PLN_AREA_N"])
```

## Working with Projection

### Assigning EPSG code to a simple feature data frame

```{r}
# To check the CSR
st_crs(mpsz)
```

```{r}
# To assign the correct EPSG code to mpsz data frame
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r}
# To check the CSR again
st_crs(mpsz3414)
```

### Transforming the projection of the preschool from wg84 to svy21.

```{r}
# To reproject preschool from one coordinate system i.e., wg84 to another i.e., svy21.
preschool3414 <- st_transform(preschool,
                              crs = 3414)
```

```{r}
# To check the CRS
st_crs(preschool3414)
```
